---
title: "3: Logistic regression"
author: "Jussi Vehviläinen"
date: "2022-11-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Assignment 3 - Logistic regression
## Bring Data to R

For the study material from UCI Machine Learning Repository, Student Performance Data (incl. Alcohol consumption) page (https://archive.ics.uci.edu/ml/datasets/Student+Performance) has been used as a source data table. 
The joined data set which will be used in the analysis, contains two student alcohol consumption data sets from the web page. The variables not used for joining the two data have been combined by averaging (including the grade variables):
  'alc_use' is the average of 'Dalc' and 'Walc'
  'high_use' is TRUE if 'alc_use' is higher than 2 and FALSE otherwise
More information about variables can be found from:
https://archive.ics.uci.edu/ml/datasets/Student+Performance

```{r}
library(tidyverse)

# Read data table to R - environment
wrangled_student_mat_por <- as.data.frame(read_csv("Data/wrangled_student_mat_por.csv"))
colnames(wrangled_student_mat_por)
```
## Select intresting variables 
For studying variables relationships to alcohol, I have selected four variables that I hypothesize to have relationship to person alcohol consumption.
Age - I think that persons over 18 use more alcohol.
romantic - I think that single persons drinks more often.
studytime - I think persons which study more weekly, drink less.
freetime - I think persons which have more free time might drink more.

```{r}
library(ggplot2)
library(GGally)

# Summary table
int_variables<-c("age","romantic","studytime","freetime","alc_use","high_use")
summary(wrangled_student_mat_por[int_variables])

# Unique values for romantic column
unique(wrangled_student_mat_por$romantic)

# Group by romantic before summarise
wrangled_student_mat_por %>% group_by(romantic) %>% summarise(count = n())

# Scatterplot
p <- ggpairs(wrangled_student_mat_por[int_variables], mapping = aes(), lower =list(combo = wrap("facethist", bins = 20))) 
p
```

Results shows that there seems to be somekind of relationship atleast between alchol and freetime, - studytime and age. POsitive correlation can be seem between alcohol use and freetime. Negative correlation between alcohol use and study time, and alcohol use and age. Intresting find out is that young persons (15-18) seems to drink more than ages 19 -21.

## Logistic regression
- Use `glm()` to fit a logistic regression model with `high_use` as the target variable and `freetime, studytime, age and romantic` as the predictors.

```{r}
m <- glm(high_use ~ freetime + studytime + age + factor(romantic), data = wrangled_student_mat_por, family = "binomial")

# Summary
summary(m)

# compute odds ratios (OR)
OR <- coef(m) %>% exp

# compute confidence intervals (CI)
CI <- confint(m) %>% exp

# print out the odds ratios with their confidence intervals
cbind(OR, CI)

```
The results shows that widest range is for variable 'romantic'. Range also contains 1 so most propably alcohol_consumption and romantic don't have relationships, more like the variables are independent of each other.
Freetime and age has OD and condidence interval  > 1, so there seems to be positive correlation between those and high_use. Studytime values are < 1 meaning negative correlation between it and high_use.

## Predictive power of the model

```{r}
# Let's select only freetime, studytime and age for next model
m <- glm(high_use ~ freetime + studytime + age, data = wrangled_student_mat_por, family = "binomial")

# Summary
summary(m)

# Prediction plots
par(mfrow=c(2,2))
plot(m)

# Probability
wrangled_student_mat_por <- mutate(wrangled_student_mat_por, probability = predict(m, type = "response"))
# Prediction
wrangled_student_mat_por <- mutate(wrangled_student_mat_por, prediction = probability > 0.5)

# define a loss function (mean prediction error)
loss_func <- function(class, prob) {
  n_wrong <- abs(class - prob) > 0.5
  mean(n_wrong)
}

# call loss_func to compute the average number of wrong predictions in the (training) data
loss_func(class = wrangled_student_mat_por$high_use, prob = 0)

```
## 10-fold cross-validation 

```{r}
library(caret)

#specify the cross-validation method
ctrl <- trainControl(method = "cv", number = 10)

#fit a regression model and use k-fold CV to evaluate performance
model <- train(factor(high_use) ~ freetime + studytime + age, data = wrangled_student_mat_por, family = "binomial", method = "glm", trControl = ctrl)
model
summary(model)

#view final model
model$finalModel

```
## Perform cross-validation to compare the performance of different logistic regression models

```{r}

# Define training control
set.seed(123)
train.control <- trainControl(method = "repeatedcv", 
                              number = 10, repeats = 3)
# Train the model
model <- train(Fertility ~., data = swiss, method = "lm",
               trControl = train.control)
# Summarize the results
print(model)
```


