---
title: '5: Dimensionality reduction techniques'
author: "Jussi Vehviläinen"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup_week5, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 5: Dimensionality reduction techniques

## Data wrangling

Data wrangling was started in last week, and script can be found from file create_human.R. There should be 195 observations and 19 variables. Most of the variables names has been shortened and couple new variables added to the original dataset as follow:
*   "GNI" = Gross National Income per capita
*   "Life.Exp" = Life expectancy at birth
*   "Edu.Exp" = Expected years of schooling 
*   "Mat.Mor" = Maternal mortality ratio
*   "Ado.Birth" = Adolescent birth rate


*   "Parli.F" = Percetange of female representatives in parliament
*   "Edu2.F" = Proportion of females with at least secondary education
*   "Edu2.M" = Proportion of males with at least secondary education
*   "Labo.F" = Proportion of females in the labour force
*   "Labo.M" " Proportion of males in the labour force

New variables
*   "Edu2.FM" = Edu2.F / Edu2.M
*   "Labo.FM" = Labo2.F / Labo2.M

There is still some editing e.g. Gender Inequality Index could be also shortened to 'GII'. Lets' start wrangling by that.


```{r data_wrangling}
# Read file to R-environment
human_data<-read.table("human_data.tsv", header=TRUE, sep=";", dec=",")


# EDIT Gender Inequality Index column
colnames(human_data)[colnames(human_data)=="Gender Inequality Index"]<-"GII"

# Libraries
library(dplyr)

# Change GNI col to numeric by mutate
human_data$GNI<-human_data %>%
  dplyr:::select(GNI) %>%
  mutate(GNI =as.numeric(GNI))


# Exclude unneeded variables
keep <- c("Country", "Edu2.FM", "Labo.FM", "Edu.Exp", "Life.Exp", "GNI", "Mat.Mor", "Ado.Birth", "Parli.F")
human_data.subset <- human_data[,keep]

# Editing previous way 'GNI' make the column to class 'dataframe'. This would be problematic in plots
# so lets' changed it back to really numeric
human_data.subset$GNI<-as.numeric(unlist(human_data.subset$GNI))

# Remove rows with all NA
human_data.subset<-human_data.subset[rowSums(is.na(human_data.subset))!= ncol(human_data.subset),]
dim(human_data.subset)

# Seems that there isn't any complete NA rows
# Lets' remove rows that contains even one NA
human_data.subset<-human_data.subset[complete.cases(as.matrix(human_data.subset)), ]
dim(human_data.subset)

# After subsetting there are 162/195 observations which haven't had any missing variables

# Next we should remove observations related to region
# to find out which, unique values of Country column should be taken

paste(unique(human_data.subset$Country),collapse=', ')

# From the print:
regions<-c("South Africa","South Asia","Latin America and the Caribbean","Europe and Central Asia","East Asia and the Pacific","World","Sub-Saharan Africa")

# Subset regions from the human_data
human_data.subset <- subset(human_data.subset, !(Country %in% regions))

# We can check by dimensions and unique values that correct rows has been removed
dim(human_data.subset)

# add country to rownames
rownames(human_data.subset) <- human_data.subset$Country

human_data.subset <- human_data.subset[,!(colnames(human_data.subset) %in% c('Country'))]

# Check dimensions, should be 155 x 8
dim(human_data.subset)

write.csv2(human_data.subset, "human_data_30112022.tsv", row.names=TRUE)
```

## Analysis

This week topic is Dimensionality reduction techniques.

### Summary of the data
```{r Summary_of_Data, echo=TRUE}
# Libraries
library(tidyverse)
library(rstatix)
library(ggpubr)
library(MASS)
library(ggplot2)
library(GGally)
library(reshape2)
library(tidyr)
library(corrplot)
library(factoextra)
library(FactoMineR)

# Let's try new package for producion of summary table and plot of the data
library(modelsummary)

# head of data
head(human_data.subset)

# Summary of the Human Data
datasummary_skim(human_data.subset)

# Summary plot
## Scatterplot
ggpairs(human_data.subset, mapping = aes(), lower =list(combo = wrap("facethist", bins = 20))) 

# Correlation plot
datasummary_correlation(human_data.subset)
```
There are differences between variables in value scale and distributions. E.g. our new variables has information combined from to column showing ratios, GNI has highest values and highest range. Overall, values difference from 0 to 123124 meaning that some sort of scaling and normalization should be done to make different variables to be comparable.

### PCA raw data

Principal component analysis (PCA) is a popular technique for analyzing large datasets containing a high number of dimensions/features per observation, increasing the interpretability of data while preserving the maximum amount of information, and enabling the visualization of multidimensional data. Formally, PCA is a statistical technique for reducing the dimensionality of a dataset. This is accomplished by linearly transforming the data into a new coordinate system where (most of) the variation in the data can be described with fewer dimensions than the initial data. (wikipedia)
```{r PCA raw data, echo=TRUE}

# perform principal component analysis (with the SVD method)
pca_human <- prcomp(human_data.subset)

# draw a biplot of the principal component representation and the original variables
biplot(pca_human, choices = 1:2)

```


From the plot, we can't see much or use it like that for good interpretation. As far, we can say that GNI seems to influence PC1.

### PCA standardized data
```{r scale_pca, echo=TRUE}

# standardize the variables
human_std <- scale(human_data.subset)

# perform principal component analysis (with the SVD method)
pca_human <- prcomp(human_std)

# draw a biplot of the principal component representation and the original variables
biplot(pca_human, choices = 1:2)

```


After scaling we can see much better different variables influence to PC1 and PC2. Edu.Exp, Life.Exp, Labb,FM and GNI influence to PC1 forming a group as well as Mat.Mor and Ado.Birth which influence is opposite and forms a group which have similar influence to PC1. Let's read a plot e.g. we can see that Niger and Sierra Leone has high Mat.Mor and Ado.Birth	and low GNI, Labo.FM, Life.Exp and Edu.Exp. This sounds realistic. And variables values in scale for e.g Japan, Korea and Czech are opposite (on the other side of mean/median).  
There could be also negative correlation with these groups, and inside group positive correlation with variables. This could be study more but we can validate this from ealier build summary plots where correlation values can be read.
Overall, we can see that these group formed from similarly influencing variables divide our countries to groups which have same level 'standard of living'.
PC2 is influenced by Padi.F and Edu2.FM which have also positive correlation to each other. It shows that countries e.g. Bolivia, Rwanda and Namibia seems to have similar level education between gender and it positivily correlates with Percetange of female representatives in parliament. As opposite, e.g Jordan has high difference between genders education level and also Percetange of female representatives in parliament level is low.


### Multiple Correspondence Analysis (MCA)
In statistics, multiple correspondence analysis (MCA) is a data analysis technique for nominal categorical data, used to detect and represent underlying structures in a data set. It does this by representing data as points in a low-dimensional Euclidean space. The procedure thus appears to be the counterpart of principal component analysis for categorical data. MCA can be viewed as an extension of simple correspondence analysis (CA) in that it is applicable to a large set of categorical variables.(wikipedia)
```{r Tea_Time, echo=TRUE}


# Bring Tea data
tea <- read.csv("https://raw.githubusercontent.com/KimmoVehkalahti/Helsinki-Open-Data-Science/master/datasets/tea.csv", stringsAsFactors = TRUE)

# Study data
View(tea)
# Dimensions
dim(tea)
# Colnames
colnames(tea)
# Summary
summary(tea)

# Summary plot
# Let's visualize part of the data
keep_columns_2<- c("price","Tea")
# Mosaic plot of couple variables
mosaicplot(table(tea[,colnames(tea) %in% keep_columns_2]), xlab='Tea', ylab='Price',main='Tea and price', col='blue')

# Age has to be factorized by bins or removed --> remove
tea<-tea[!(colnames(tea) %in% 'age')]

# Multiple Correspondence Analysis (MCA)
# set to vizualize also the correlation between variables and MCA principal dimension
res.mca<- MCA(tea, graph = TRUE)
# With so many variables the printed MCAfactor plots becomes harder to read. We can see that e.g tea shop and older age seems to have positive influence to dim 2 where as student positivily correlates with age group 15-24 and have negative influence to dim 2.
#But by selecting less variables we might make better conclusions.
# Let's use only 12 first and check if can we get better plots for inspecting.
res.mca<- MCA(tea[,1:12], graph = TRUE)
# Extract and save results for later use
var <- get_mca_var(res.mca)
var
# Coordinates
head(var$coord)
# Cos2: quality on the factore map
head(var$cos2)
# Contributions to the principal components
head(var$contrib)     

# visualize MCA - many possibilites
# MCA factor map
plot(res.mca, invisible=c("ind"), graph.type = "classic")


# Eigenvalues/Variances
# get values
eig.val <- get_eigenvalue(res.mca)
head(eig.val)
# Build plot
# Shows how much first ten dimension of data explains it"
fviz_screeplot(res.mca, addlabels = TRUE, ylim = c(0, 45))


# Biplot
# The plot above shows a global pattern within the data. Rows (individuals) are represented by blue points and columns (variable categories) by red triangles.

# The distance between any row points or column points gives a measure of their similarity (or dissimilarity). Row points with similar profile are closed on the factor map. The same holds true for column points.
fviz_mca_biplot(res.mca, 
               repel = TRUE,
               ggtheme = theme_minimal())



# Cos2
# The quality of the representation is called the squared cosine (cos2), which measures the degree of association between variable categories and a particular axis. 

# Color by cos2 values: quality on the factor map
fviz_mca_var(res.mca, col.var = "cos2",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), 
             repel = TRUE, # Avoid text overlapping
             ggtheme = theme_minimal())

# Contribution of variable categories to the dimensions 1-2
fviz_contrib(res.mca, choice = "var", axes = 1:2, top = 15)
# The red dashed line on the graph above indicates the expected average value, If the contributions were uniform.
fviz_mca_var(res.mca, col.var = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), 
             repel = TRUE, # avoid text overlapping (slow)
             ggtheme = theme_minimal()
             )
# The plot above gives an idea of what pole of the dimensions the categories are actually contributing to. 
