---
title: '5: Dimensionality reduction techniques'
author: "Jussi Vehviläinen"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 5: Dimensionality reduction techniques

## Data wrangling

Data wrangling was started in last week, and script can be found from file create_human.R. There should be 195 observations and 19 variables. Most of the variables names has been shortened and couple new variables added to the original dataset as follow:
*   "GNI" = Gross National Income per capita
*   "Life.Exp" = Life expectancy at birth
*   "Edu.Exp" = Expected years of schooling 
*   "Mat.Mor" = Maternal mortality ratio
*   "Ado.Birth" = Adolescent birth rate


*   "Parli.F" = Percetange of female representatives in parliament
*   "Edu2.F" = Proportion of females with at least secondary education
*   "Edu2.M" = Proportion of males with at least secondary education
*   "Labo.F" = Proportion of females in the labour force
*   "Labo.M" " Proportion of males in the labour force

New variables
*   "Edu2.FM" = Edu2.F / Edu2.M
*   "Labo.FM" = Labo2.F / Labo2.M

There is still some editing e.g. Gender Inequality Index could be also shortened to 'GII'. Lets' start wrangling by that.


```{r data_wrangling}
# Read file to R-environment
human_data<-read.table("human_data.tsv", header=TRUE, sep=";", dec=",")

# EDIT Gender Inequality Index column
colnames(human_data)[colnames(human_data)=="Gender Inequality Index"]<-"GII"

# Libraries
library(dplyr)

# Change GNI col to numeric by mutate
human_data %>%
  select(GNI) %>%
  mutate(GNI =as.numeric(GNI))

# Exclude unneeded variables
keep <- c("Country", "Edu2.FM", "Labo.FM", "Edu.Exp", "Life.Exp", "GNI", "Mat.Mor", "Ado.Birth", "Parli.F")
human_data.subset <- human_data[,keep]

# Remove rows with all NA
human_data.subset<-human_data.subset[rowSums(is.na(human_data.subset))!= ncol(human_data.subset),]
dim(human_data.subset)

# Seems that there isn't any complete NA rows
# Lets' remove rows that contains even one NA
human_data.subset<-human_data.subset[complete.cases(human_data.subset), ]
dim(human_data.subset)

# After subsetting there are 162/195 observations which haven't had any missing variables

# Next we should remove observations related to region
# to find out which, unique values of Country column should be taken

unique(human_data.subset$Country)

# From the print:
regions<-c("South Africa","South Asia","Latin America and the Caribbean","Europe and Central Asia","East Asia and the Pacific","World","Sub-Saharan Africa")

# Subset regions from the human_data
human_data.subset <- subset(human_data.subset, !(Country %in% regions))

# We can check by dimensions and unique values that correct rows has been removed
dim(human_data.subset)
unique(human_data.subset$Country)

# add country to rownames
rownames(human_data.subset) <- human_data.subset$Country

human_data.subset <- human_data.subset[,!(colnames(human_data.subset) %in% c('Country'))]

# Check dimensions, should be 155 x 8
dim(human_data.subset)

write.csv2(human_data.subset, "human_data_30112022.tsv", row.names=TRUE)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
