---
title: '6: Analysis of longitudinal data'
author: "Jussi Vehviläinen"
date: "`r Sys.Date()`"
output: html_document
---

```{r week6, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 6: Analysis of longitudinal data

*   Show as much of the relevant raw data as possible rather than only data summaries;
*   Highlight aggregate patterns of potential scientific interest;
*   Identify both cross-sectional and longitudinal patterns;
*   Try to make the identification of unusual individuals or unusual observations simple.


## RATS Data analysis

For the longitudinal data, is typical that the response variable (and possibly some or all of the explanatory variables) is observed more than once on each individual in the study and so could be labelled multivariate.
RATS Data set contains three groups with each containing 16 subject. Rats mass changes has been followed 64 days and mass has been weighted in seven days period.

### Bring data
```{r cars}
# Bring data to R and check summaries
ratsl=read.table('ratsl.txt', header=TRUE)
str(ratsl)
ratsl$Group = factor(ratsl$Group)
ratsl$ID = factor(ratsl$ID)
str(ratsl)
```

### Graphical Displays of Longitudinal Data
```{r ratsl, echo=TRUE}
# Needed R-libraries
library(dplyr);library(tidyr);library(ggplot2);library(rstatix)

# Draw the plot
ggplot(ratsl, aes(x = time_variable , y = rats, linetype = ID, color= ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(ratsl$rats), max(ratsl$rats)))
## There seems to quite wide range between different group rats masses. Highest masses are in group 2 masses and lowest in group 1. Base level changes between 250 - 550 g. To make groups more comparable, data should be standardized.

# Standardize the data set
ratsl <- ratsl %>%
  group_by(time_variable) %>%
  mutate( stdratsl = (rats - mean(rats))/sd(rats) ) %>%
  ungroup()

# Check and study edited data frame
str(ratsl)
glimpse(ratsl)
# Build new ggplot by using standardized values - mean profiles
ggplot(ratsl, aes(x = time_variable, y = stdratsl, linetype = ID, color = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") +
  scale_y_continuous(limits = c(min(ratsl$stdratsl), max(ratsl$stdratsl)))
## After standardization we can see that standardized masses varies between ~ - 1.25 - 1.7. 

# Summary data with mean and standard error of bprs by treatment and week 
ratss <- ratsl %>%
  group_by(time_variable, Group) %>%
  summarise(mean = mean(rats), se = sd(rats)) %>%
  ungroup()

# Glimpse the data
glimpse(ratss)

# Plot the mean profiles
ggplot(ratss, aes(x = time_variable, y = mean, linetype = Group, shape=Group, color=Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.8,0.8)) +
  scale_y_continuous(name = "mean(ratss) +/- se(ratss)")
## From the plot, it's easy to see that mean in group 1 are much lower than in groups 2 and 3 which error bars overlap each other more when time goes on. By excluding error bar we can study more difference between group 2 and 3.


# Plot the mean profiles without error bar
ggplot(ratss, aes(x = time_variable, y = mean, linetype = Group, shape=Group, color=Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  #geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.8,0.8)) +
  scale_y_continuous(name = "mean(ratss) +/- se(ratss)")
## Now we can see that group2 has higher change in mass in later days than group 2, resulting in a smaller range between groups.

# Create a summary data by Group and subject with mean as the summary variable (ignoring baseline time variable 1)
ratsl8s <- ratsl %>%
  filter(time_variable > 1) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(rats) ) %>%
  ungroup()

# Glimpse the data
glimpse(ratsl8s)

# Draw a boxplot of the mean versus treatment
ggplot(ratsl8s, aes(x = Group, y = mean, group=Group, color=Group)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(rats), time variable 1-64")
## Every group seems to contain one 'outlier'.

# Lets' add names to 'outliers' so that those can be easily locate and delete
## Function for finding those
is_outlier <- function(x) {
  return(x < quantile(x, 0.25) - 1.5 * IQR(x) | x > quantile(x, 0.75) + 1.5 * IQR(x))
}

ratsl8s %>%
  group_by(Group) %>%
  mutate(outlier = ifelse(is_outlier(mean), mean, as.numeric(NA))) %>%
  ggplot(., aes(x = Group, y = mean, group=Group, color=Group)) +
    geom_boxplot() +
    geom_text(aes(label = outlier), na.rm = TRUE, hjust = -0.3) +
    stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
    scale_y_continuous(name = "mean(rats), time variable 1-64")
## From the plot we can read that 238.9 , 594 & 495.2 are 'outliers'.

# Create a new data by filtering 'the outlier' 
ratsl8s_wo_O <- subset(ratsl8s,!(mean %in% c(238.9,594,495.2)))
## Let's use same function with ggplot so we can be sure that there isn't any outliers
ratsl8s_wo_O %>%
  group_by(Group) %>%
  mutate(outlier = ifelse(is_outlier(mean), mean, as.numeric(NA))) %>%
  ggplot(., aes(x = Group, y = mean, group=Group, color=Group)) +
    geom_boxplot() +
    geom_text(aes(label = outlier), na.rm = TRUE, hjust = -0.3) +
    stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
    scale_y_continuous(name = "mean(rats), time variable 1-64")
## Now the box plots seems better so we can make statistical test

# Because there are multiple groups, pairwise t-tests and bonferroni p.adjust.method from R-package rstatix could be used.
pairwise_t_test(mean ~ Group, data=ratsl8s_wo_O, p.adjust.method = "bonferroni")
## From the result table we can see that there are strongly significant differences between all study groups.

# Add the baseline from the original data as a new variable to the summary data
## First read original data to R.
rats <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", header = TRUE, sep = '\t')
## Mutate Baseline WD1 to summary table
ratsl8s_2 <- ratsl8s %>%
  mutate(baseline = rats$WD1)
## Check
str(ratsl8s_2)

# Fit the linear model with the mean as the response 
fit <- lm(mean ~ baseline + Group, data = ratsl8s_2)
summary(fit)
## From the linear model summary we can see that baseline seems to be best and only significant predictor for the mean.

```

## BPRS Data analyses

The brief psychiatric rating scale (BPRS) data set contains 40 male subjects which were randomly assigned to one of two treatment groups and each subject was rated on the BPRS measured before treatment began (week 0) and then at weekly intervals for eight weeks. The BPRS assesses the level of 18 symptom constructs such as hostility, suspiciousness, hallucinations and grandiosity; each of these is rated from one (not present) to seven (extremely severe). The scale is used to evaluate patients suspected of having schizophrenia.

### Bring data
```{r bprsl}
# Bring data to R
bprsl=read.table('BPRSL.txt', header=TRUE)
# Check and study datas' structures.
str(bprsl)
bprsl$treatment = factor(bprsl$treatment)
bprsl$subject <- factor(bprsl$subject)
# Check and study datas' structures after changes.
str(bprsl)
```

### Analysis of Longitudinal Data II: Linear Mixed Effects Models for Normal Response Variables
```{r BPRS, echo=TRUE}
# Needed R-libraries
library(dplyr);library(tidyr);library(ggplot2);library(rstatix);library(lme4)

# Plot the RATSL data
ggplot(bprsl, aes(x = week, y = bprs, color = subject)) +
  geom_line(aes(linetype = treatment)) +
  scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 8, 1)) +
  scale_y_continuous(name = "BPRS") +
  theme(legend.position = "top")
## From the one single plot, differences between subjects and treatments is hard to read. There seem to be negative correlation between BPRS and time, but it should be studied more.

# Create a regression model bprsl_reg
bprsl_reg <- "Regression model here!"
# Fit the linear model with the mean as the response 
bprsl_reg <- lm(bprs ~ treatment + week, data = bprsl)
summary(bprsl_reg)
# Week aka time seems to be better predictor than treatment

# Lets' try another model
## Create a random intercept model which takes account random effect caused by subject aka patient.
bprsl_ref <- lmer(bprs ~ treatment + week + (1 | subject), data = bprsl, REML = FALSE)
summary(bprsl_ref)

# Also time - subject combination might cause random effect
# create a random intercept and random slope model
bprsl_ref1 <- lmer(bprs ~ treatment + week + (week | subject), data = bprsl, REML = FALSE)
summary(bprsl_ref1)

# Two models results can be compared by performing an ANOVA test on the two models
anova(bprsl_ref, bprsl_ref1)
## From the results we can say that later model (bprsl_ref1) seems to be better one

# To learn more about random effect cross variable 'treatment * week' can be added to the model
## Create a random intercept and random slope model with the interaction
bprsl_ref2 <- lmer(bprs ~ treatment * week + (week | subject), data = bprsl, REML = FALSE)
summary(bprsl_ref2)

# Compare two latest model
## Perform an ANOVA test on the two models
anova(bprsl_ref2, bprsl_ref1)
## Now the model bprsl_ref2 seems to be better for this dataset.

# Draw the plot of bprsl with the observed Weight values
ggplot(bprsl, aes(x = week, y = bprs, colour = subject)) +
  geom_line(aes(linetype = treatment)) +
  scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 8, 1)) +
  scale_y_continuous(name = "Observed BPRS") +
  theme(legend.position = "top")
## To clarify the negative correlation, fitted values might give better visualization.

# Create a vector of the fitted values
Fitted <- fitted(bprsl_ref2)

# Create a new column fitted to bprsl
bprsl$Fitted_bprs = Fitted

# Draw the plot of bprsl with the Fitted values of weight
ggplot(bprsl, aes(x = week, y = Fitted_bprs, colour = subject)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  geom_line(aes(linetype = treatment)) +
  scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 8, 1)) +
  scale_y_continuous(name = "Fitted BPRS") +
  theme(legend.position = "top")
## After fitting the model, we can clearly see that there are negative correlation with weeks and BPRS in both treatment groups
```

