---
title: '6: Analysis of longitudinal data'
author: "Jussi Vehviläinen"
date: "`r Sys.Date()`"
output: html_document
---

```{r week6, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 6: Analysis of longitudinal data

*   Show as much of the relevant raw data as possible rather than only data summaries;
*   Highlight aggregate patterns of potential scientific interest;
*   Identify both cross-sectional and longitudinal patterns;
*   Try to make the identification of unusual individuals or unusual observations simple.


## RATS Data analysis

### Bring data
```{r cars}
# Bring data to R and check summaries
ratsl=read.table('ratsl.txt', header=TRUE)
str(ratsl)
ratsl$Group = factor(ratsl$Group)
ratsl$ID = factor(ratsl$ID)
str(ratsl)
```

### Graphical Displays of Longitudinal Data
```{r ratsl, echo=TRUE}
# Needed R-libraries
library(dplyr);library(tidyr);library(ggplot2);library(rstatix)

# Draw the plot
ggplot(ratsl, aes(x = time_variable , y = rats, linetype = ID, color= ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(ratsl$rats), max(ratsl$rats)))

# Standardize the data to see how it effects to the plots

ratsl <- ratsl %>%
  group_by(time_variable) %>%
  mutate( stdratsl = (rats - mean(rats))/sd(rats) ) %>%
  ungroup()

# Check edited data frame
str(ratsl)
glimpse(ratsl)
# Build new ggplot by using standardised values - mean profiles
ggplot(ratsl, aes(x = time_variable, y = stdratsl, linetype = ID, color = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") +
  scale_y_continuous(limits = c(min(ratsl$stdratsl), max(ratsl$stdratsl)))


# Summary data with mean and standard error of bprs by treatment and week 
ratss <- ratsl %>%
  group_by(Group, time_variable) %>%
  summarise( mean = rats, se = rats) %>%
  ungroup()

# Glimpse the data
glimpse(ratss)

# Plot the mean profiles
ggplot(ratss, aes(x = time_variable, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.8,0.8)) +
  scale_y_continuous(name = "mean(ratss) +/- se(ratss)")

# Create a summary data by Group and subject with mean as the summary variable (ignoring baseline time variable 1)
ratsl8s <- ratsl %>%
  filter(time_variable > 1) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(rats) ) %>%
  ungroup()

# Glimpse the data
glimpse(ratsl8s)

# Draw a boxplot of the mean versus treatment
ggplot(ratsl8s, aes(x = Group, y = mean, group=Group, color=Group)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(rats), time variable 1-64")

# Lets' add names to outliers so that those can be easily locate and delete
## Function for finding those
is_outlier <- function(x) {
  return(x < quantile(x, 0.25) - 1.5 * IQR(x) | x > quantile(x, 0.75) + 1.5 * IQR(x))
}

ratsl8s %>%
  group_by(Group) %>%
  mutate(outlier = ifelse(is_outlier(mean), mean, as.numeric(NA))) %>%
  ggplot(., aes(x = Group, y = mean, group=Group, color=Group)) +
    geom_boxplot() +
    geom_text(aes(label = outlier), na.rm = TRUE, hjust = -0.3) +
    stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
    scale_y_continuous(name = "mean(rats), time variable 1-64")

# Create a new data by filtering the outlier 
## From the plot we can read that 238.9 , 594 & 495.2 are outliers
ratsl8s_wo_O <- subset(ratsl8s,!(mean %in% c(238.9,594,495.2)))

# Let's use same function with ggplot so we can be sure that there isn't any outliers
ratsl8s_wo_O %>%
  group_by(Group) %>%
  mutate(outlier = ifelse(is_outlier(mean), mean, as.numeric(NA))) %>%
  ggplot(., aes(x = Group, y = mean, group=Group, color=Group)) +
    geom_boxplot() +
    geom_text(aes(label = outlier), na.rm = TRUE, hjust = -0.3) +
    stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
    scale_y_continuous(name = "mean(rats), time variable 1-64")

# Now the boxplots seems better so we can make statistical test
## Because we multiple groups, we will use pairwise t-tests and bonferroni p.adjust.method
pairwise_t_test(mean ~ Group, data=ratsl8s_wo_O, p.adjust.method = "bonferroni")
# From the result table we can see that there are significant differences between all groups


# Add the baseline from the original data as a new variable to the summary data
rats <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", header = TRUE, sep = '\t')

ratsl8s_2 <- ratsl8s %>%
  mutate(baseline = rats$WD1)

# Fit the linear model with the mean as the response 
fit <- lm(mean ~ baseline + Group, data = ratsl8s_2)
summary(fit)
# From the linear model summary we can see that baseline seems to best predictor for the mean.

```

## BPRS Data analyses

A number of graphical displays which can be useful in the preliminary assessment of longitudinal data from clinical trials will now be illustrated using the data shown below taken from Davis (2002). Here 40 male subjects were randomly assigned to one of two treatment groups and each subject was rated on the brief psychiatric rating scale (BPRS) measured before treatment began (week 0) and then at weekly intervals for eight weeks. The BPRS assesses the level of 18 symptom constructs such as hostility, suspiciousness, halluci- nations and grandiosity; each of these is rated from one (not present) to seven (extremely severe). The scale is used to evaluate patients suspected of having schizophrenia

### Bring data
```{r bprsl}
# Bring data to R
bprsl=read.table('BPRSL.txt', header=TRUE)
str(bprsl)
bprsl$treatment = factor(bprsl$treatment)
bprsl$subject <- factor(bprsl$subject)

str(bprsl)

```

### Analysis of Longitudinal Data II: Linear Mixed Effects Models for Normal Response Variables
```{r BPRS, echo=TRUE}
# Needed R-libraries
library(dplyr);library(tidyr);library(ggplot2);library(rstatix);library(lme4)

# Plot the RATSL data
ggplot(bprsl, aes(x = week, y = bprs, color = subject)) +
  geom_line(aes(linetype = treatment)) +
  scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 8, 1)) +
  scale_y_continuous(name = "BPRS") +
  theme(legend.position = "top")

# create a regression model bprsl_reg
bprsl_reg <- "Regression model here!"
# Fit the linear model with the mean as the response 
bprsl_reg <- lm(bprs ~ treatment + week, data = bprsl)
summary(bprsl_reg)
# Week aka time seems to be better predictor than treatment

# Lets' try another model
# Create a random intercept model which takes account random effect caused by patient aka subject
bprsl_ref <- lmer(bprs ~ treatment + week + (1 | subject), data = bprsl, REML = FALSE)
summary(bprsl_ref)

# Also time - subject combination might cause random effect
# create a random intercept and random slope model
bprsl_ref1 <- lmer(bprs ~ treatment + week + (week | subject), data = bprsl, REML = FALSE)
summary(bprsl_ref1)

# Two models results can be compared
# perform an ANOVA test on the two models
anova(bprsl_ref, bprsl_ref1)
# From the results we can say that later model seems to be better one

# create a random intercept and random slope model with the interaction
bprsl_ref2 <- lmer(bprs ~ treatment * week + (week | subject), data = bprsl, REML = FALSE)
summary(bprsl_ref2)

# Compare two latest model
# perform an ANOVA test on the two models
anova(bprsl_ref2, bprsl_ref1)

# draw the plot of bprsl with the observed Weight values
ggplot(bprsl, aes(x = week, y = bprs, colour = subject)) +
  geom_line(aes(linetype = treatment)) +
  scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 8, 1)) +
  scale_y_continuous(name = "Observed BPRS") +
  theme(legend.position = "top")

# Create a vector of the fitted values
Fitted <- fitted(bprsl_ref2)

# Create a new column fitted to bprsl
bprsl$Fitted_bprs = Fitted

# draw the plot of bprsl with the Fitted values of weight
ggplot(bprsl, aes(x = week, y = Fitted_bprs, colour = subject)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  geom_line(aes(linetype = treatment)) +
  scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 8, 1)) +
  scale_y_continuous(name = "Observed BPRS") +
  theme(legend.position = "top")

# After fitting the model, we can clearly see that there are negative correlation with weeks and BPRS in both treatment groups
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
