---
title: "4: Clustering and classification"
author: "Jussi Vehviläinen"
date: "2022-11-24"
output: html_document
---

```{r setup_week4, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Assignment 4: Clustering and classification

This assignment topic is clustering and classification. As basic, clustering means that in the data some points/observations are in space so near each other compared to other points/observations that those form a 'own cluster'.
There are multiple different clustering methods to find those clusters from the data. One of the most common methods is k-means clustering, others worth naming are hierarchical clustering methods which gives dendrograms as a output.
If the clustering can be done successfully, new observations can be tried to classify to those clusters.

To clarify this topic, Boston dataset from MASS - package will be used as a demonstration dataset.

The Boston Dataset

The Boston Housing Dataset is a derived from information collected by the U.S. Census Service concerning housing in the area of Boston MA. The Boston data frame has 506 rows and 14 columns. The following describes the dataset columns:

*   CRIM - per capita crime rate by town
*   ZN - proportion of residential land zoned for lots over 25,000 sq.ft.
*   INDUS - proportion of non-retail business acres per town.
*   CHAS - Charles River dummy variable (1 if tract bounds river; 0 otherwise)
*   NOX - nitric oxides concentration (parts per 10 million)
*   RM - average number of rooms per dwelling
*   AGE - proportion of owner-occupied units built prior to 1940
*   DIS - weighted distances to five Boston employment centres*
*   RAD - index of accessibility to radial highways
*   TAX - full-value property-tax rate per $10,000
*   PTRATIO - pupil-teacher ratio by town
*   B - 1000(Bk - 0.63)^2 where Bk is the proportion of blacks by town
*   LSTAT - % lower status of the population
 *  MEDV - Median value of owner-occupied homes in $1000's

## Summary and graphical overview of the data
```{r Summary, echo=FALSE}
# Bring needed libraries to R-environment
library(MASS)
library(ggplot2)
library(GGally)
library(reshape2)
library(tidyr)
library(corrplot)

# Check head of Boston table
head(Boston)

# Check dimensions
dim(Boston)

# Check summary of data
summary(Boston)

# Visualize data for inspecting
## Scatterplot
ggpairs(Boston, mapping = aes(), lower =list(combo = wrap("facethist", bins = 20))) 

## All variables against crim rate
bosmelt <- melt(Boston, id="crim")
ggplot(bosmelt, aes(x=value, y=crim))+
  facet_wrap(~variable, scales="free")+
  geom_point()

# Correlation matrix and visualization
## Matrix
corrmatrix <- cor(Boston)
corrmatrix
## Visualization
corrplot(corrmatrix, method="circle", cl.pos = "b", tl.pos = "d", tl.cex = 0.6, type = "upper")
```
There are 506 different observations for 14 variables whivh summaries informations concerning housing in the area of Boston Mass. By using this dataset, we can try to find out different relationships between variables. E.g. per capita crime rate  relationship to other variables is very interesting. If we could find out variables which has the most effect to it, could in future try to effect these variables and in advance reduce develop of crime rate when planning new residential areas.
From the summary table, we can see that not all information from variables are easy or wise to use as a predictors in same model as those are now. E.g. "proportion of residential land zoned for lots over 25,000 sq.ft." (zn) - variable observations are unevenly distributed where as "pupil-teacher ratio by town" (ptratio) - variable observations are much more evenly distributed.
But interesting is that when we plot other variables against crime rate, we can see that there could be correlation even that the distribution wouldn't be so clear.
From the correlation matrix,  we can see that there are some kind positive or negative correlation with per capita crime rate and other variables. Six variables have negative correlation and seven have positive. Most less negative correlation is with "Charles River dummy variable (1 if tract bounds river; 0 otherwise)" (chas) and overall negative correlate factors impact seems to be low. Positively correlate factors impact seems to be much higher and "index of accessibility to radial highways" (rad) -  and "full-value property-tax rate per $10,000" (tax) seems to have the most positive correlation  to per capita crime rate. Same results are visualized in the plot where conclusions can be made more easily.

## Standardize the dataset
To get more clear understanding about the data and make it usable in prediction model, it has to be standardized. To do that, we scale and center the columns of a numeric matrix. Centering is done by subtracting the column means (omitting NAs) of x from their corresponding columns. Scaling is done by dividing the (centered) columns of x by their standard deviations.  

```{r Stand, echo=FALSE}
# Check head of original data to compare scaled data
data("Boston")
head(Boston)
# center and standardize variables, also set output to be data.frame
boston_scaled <- as.data.frame(scale(Boston, center = TRUE, scale = TRUE))
head(boston_scaled)
# Set 'crim' to numeric variable
boston_scaled$crim <- as.numeric(boston_scaled$crim)
# Summary of crim variable
summary(boston_scaled[,'crim'])

# class of the boston_scaled object
class(boston_scaled)

# Inspect how values has changed 
# Check summary of data
summary(boston_scaled)

# Visualize data for inspecting
## Scatterplot
ggpairs(boston_scaled, mapping = aes(), lower =list(combo = wrap("facethist", bins = 20))) 

## All variables against crim rate
bosmelt <- melt(boston_scaled, id="crim")
ggplot(bosmelt, aes(x=value, y=crim))+
  facet_wrap(~variable, scales="free")+
  geom_point()

# Correlation matrix and visualization
## Matrix
corrmatrix <- cor(boston_scaled)
corrmatrix
## Visualization
corrplot(corrmatrix, method="circle", cl.pos = "b", tl.pos = "d", tl.cex = 0.6, type = "upper")
```
After scaling we can see that values across different variables are much more near each others. From the correlation matrix, we see that we got same results as with raw values. 

```{r Build_crime, echo=FALSE}
# create a quantile vector of crim and print it
bins <- quantile(boston_scaled[['crim']])
bins

# create a categorical variable 'crime'
crime <- cut(boston_scaled[['crim']], breaks = bins, include.lowest = TRUE)

# look at the table of the new factor crime
head(boston_scaled)

# remove original crim from the dataset
boston_scaled <- dplyr::select(boston_scaled, -crim)

# add the new categorical value to scaled data
boston_scaled <- data.frame(boston_scaled, crime)

# Set 'crime' to factor variable, and give relevant levels
boston_scaled$crime <- factor(boston_scaled$crime, levels = c("low", "med_low", "med_high", "high"))
```


## Divide the dataset to train and test sets

* Data will be divided 80/20 to train and test sets

```{r Train_Test, echo=TRUE}
# Bring Boston data table from github to R

boston_scaled <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/Helsinki-Open-Data-Science/master/datasets/boston_scaled.txt", sep=",", header = T)

# Change 'crime' to factor variable with relevant levels
boston_scaled$crime <- factor(boston_scaled$crime, levels = c("low", "med_low", "med_high", "high"))

# Build up test and train datasets
# number of rows in the Boston dataset 
n <- nrow(boston_scaled)

# choose randomly 80% of the rows
ind <- sample(n,  size = n * 0.8)

# create train set
train <- boston_scaled[ind,]
dim(train)
# create test set 
test <- boston_scaled[-ind,]
dim(test)

# save the correct classes from test data
correct_classes <- test$crime

# remove the crime variable from test data
test <- dplyr::select(test, -crime)
head(test)
```
Now we have build up test and training data by selecting randomly first 80 % off data as a training set and using 20 % as a test set. Both data set will be used in next part. 

## Fit the linear discriminant analysis on the train set

* Crime rate as the target variable and all the other variables in the dataset as predictor variables.

Now we use training set in linear discriminat analysis. The function tries hard to detect if the within-class covariance matrix is singular. If any variable has within-group variance less than 'tolerance to decide'^2 it will stop and report the variable as constant. As a return function gives:

*prior - the prior probabilities used.

*means - the group means.

*scaling - a matrix which transforms observations to discriminant functions, normalized so that within groups covariance matrix is spherical.

*svd - the singular values, which give the ratio of the between- and within-group standard deviations on the linear discriminant variables. Their squares are the canonical F-statistics.

*N - The number of observations used.

*call	- The (matched) function call.

```{r Modelli, echo=TRUE}
# linear discriminant analysis
lda.fit <- lda(crime ~., data = train)

# print the lda.fit object
lda.fit

# the function for lda biplot arrows
lda.arrows <- function(x, myscale = 1, arrow_heads = 0.1, color = "red", tex = 0.75, choices = c(1,2)){
  heads <- coef(x)
  arrows(x0 = 0, y0 = 0, 
         x1 = myscale * heads[,choices[1]], 
         y1 = myscale * heads[,choices[2]], col=color, length = arrow_heads)
  text(myscale * heads[,choices], labels = row.names(heads), 
       cex = tex, col=color, pos=3)
}

# target classes as numeric
classes <- as.numeric(train$crime)

# plot the lda results
plot(lda.fit, dimen = 2) +
  lda.arrows(lda.fit, myscale = 1)
```
From the results plot we can see that rad seems to most driving variable in our training set for explaining LD1. And from the result object we can see that it has most positive coefficient 3.62727 to LD1 whereas second most effective variable seems to be nox with only 0.307. For the LD2, zn has highest positive coefficient 0.624225492.
## Predict the classes with the LDA model on the test data

```{r LDA Model, echo=FALSE}
# predict classes with test data
lda.pred <- predict(lda.fit, newdata = test)
lda.pred

# Define a loss function (mean prediction error)
calc_class_err <- function(actual, predicted) {
  mean(actual != predicted)
}

# call loss_func to compute the average number of wrong predictions in the (training) data
calc_class_err(actual = correct_classes, predicted = lda.pred$class)

# cross tabulate the results
table(correct = correct_classes, predicted = lda.pred$class)
```
The results shows as that the high and low variable groups are grouped much better to rigth classes compared to med_high and med_low groups. Overall our error rate is still around 33 % depending how observations are divided to train and test data. This is pretty high and could be better.

## Reload the Boston dataset and standardize the dataset 
To get better error rate we can try to do standardise our data set by using distances.
* Scale all variables to have mean = 0 and standard deviation = 1
```{r Scale, echo=FALSE}
# Bring data to R
library(MASS)
library(dplyr)
data(Boston)
# Standaridise data
standardised_boston <- as.data.frame(Boston %>%mutate_all(~(scale(.) %>% as.vector)))

# Calculate distances

## Euclidean distance matrix
dist_eu <- dist(standardised_boston, method = "euclidean")

# look at the summary of the distances
summary(dist_eu)

# manhattan distance matrix
dist_man <- dist(standardised_boston, method = "manhattan")

# look at the summary of the distances
summary(dist_man)

set.seed(13)

# K-means clustering, center 4
km <- kmeans(Boston, centers = 4)
# plot the Boston dataset with clusters
pairs(Boston, col = km$cluster)

# Summary of clusters
summary(km$cluster)

# K-means clustering, center 3
km <- kmeans(Boston, centers = 3)
# plot the Boston dataset with clusters
pairs(Boston[1:5], col = km$cluster)
pairs(Boston[6:10], col = km$cluster)
pairs(Boston[11:ncol(Boston)], col = km$cluster)

# Summary of clusters
summary(km$cluster)

# K-means clustering, center 2
km <- kmeans(Boston, centers = 2)
# plot the Boston dataset with clusters
pairs(Boston[1:5], col = km$cluster)
pairs(Boston[6:10], col = km$cluster)
pairs(Boston[11:ncol(Boston)], col = km$cluster)

# Summary of clusters
summary(km$cluster)

# K-means clustering, center 2
km <- kmeans(Boston, centers = 5)
# plot the Boston dataset with clusters
pairs(Boston[1:5], col = km$cluster)
pairs(Boston[6:10], col = km$cluster)
pairs(Boston[11:ncol(Boston)], col = km$cluster)

# Summary of clusters
summary(km$cluster)

# K-means clustering, center 2
km <- kmeans(Boston, centers = 6)
# plot the Boston dataset with clusters
pairs(Boston[1:5], col = km$cluster)
pairs(Boston[6:10], col = km$cluster)
pairs(Boston[11:ncol(Boston)], col = km$cluster)

# Summary of clusters
summary(km$cluster)

```

From the summaries, we can see that distances between used method differs quite much qiven longer distances by Manhattan method.
Euclidean distance is the shortest path between source and destination. Manhattan distance is sum of all the real distances between source and destination.  Two clusters seems to be best.
By using center 4, number of cluster range between 1-4, and 2 seems to be the most common.
When studying by setting higher value for center, 4 cluster seems to be most common choice. So, it seems to be hard to use summary as a only way to select good number of clusters. But from the plots we can see that 4 clusters could be presented in the data, but even that atleast 2 and 3 clusters should be also tried.

## Bonus: k-means on the original Boston data with 3 clusters

```{r Bonus, echo=FALSE}
data(Boston)
# Standaridise data
standardised_boston <- as.data.frame(Boston %>%mutate_all(~(scale(.) %>% as.vector)))

# K-means clustering, center 3
km <- kmeans(standardised_boston, centers = 3)
# plot the Boston dataset with clusters
pairs(standardised_boston[1:5], col = km$cluster)
pairs(standardised_boston[6:10], col = km$cluster)
pairs(standardised_boston[11:ncol(standardised_boston)], col = km$cluster)

# Summary of clusters
summary(km$cluster)

standardised_boston$N_clusters <- factor(km$cluster)

# Build up test and train datasets
# choose randomly 80% of the rows
ind <- sample(nrow(standardised_boston),  size = n * 0.8)

# create train set
train <- standardised_boston[ind,]
dim(train)
# create test set 
test <- standardised_boston[-ind,]
dim(test)

# save the 'correct' clusters from test data
correct_classes <- test$N_clusters

# remove the N_clusters variable from test data
test <- dplyr::select(test, -N_clusters)
head(test)

# linear discriminant analysis
lda.fit <- lda(N_clusters ~., data = train)

# print the lda.fit object
lda.fit

# the function for lda biplot arrows
lda.arrows <- function(x, myscale = 1, arrow_heads = 0.1, color = "red", tex = 0.75, choices = c(1,2)){
  heads <- coef(x)
  arrows(x0 = 0, y0 = 0, 
         x1 = myscale * heads[,choices[1]], 
         y1 = myscale * heads[,choices[2]], col=color, length = arrow_heads)
  text(myscale * heads[,choices], labels = row.names(heads), 
       cex = tex, col=color, pos=3)
}

# target classes as numeric
classes <- as.numeric(train$N_clusters)

# plot the lda results
plot(lda.fit, dimen = 2) +
  lda.arrows(lda.fit, myscale = 1)

```
Plot of the LDA results shows that tax and rad has most effect to LD1 and age for LD2, and even so are the most influential linear separators for the clusters.

## SuperBonus 
 
```{r SuperBonus, echo=FALSE}


# Add crime factor to data set scaled in last code chunk

# create a quantile vector of crim and print it
bins <- quantile(standardised_boston[['crim']])
bins

# create a categorical variable 'crime'
crime <- cut(standardised_boston[['crim']], breaks = bins, include.lowest = TRUE)

# remove original crim from the dataset
standardised_boston <- dplyr::select(standardised_boston, -crim)

# add the new categorical value to scaled data
standardised_boston <- data.frame(standardised_boston, crime)

# Set crime to character fow a while 
standardised_boston$crime<-as.character(standardised_boston$crime)

# Set 'crime' to factor variable, and give relevant levels
standardised_boston$crime[standardised_boston$crime=="[-0.419,-0.411]"] <- "low"
standardised_boston$crime[standardised_boston$crime=="(-0.411,-0.39]"] <- "med_low"
standardised_boston$crime[standardised_boston$crime=="(-0.39,0.00739]"] <- "med_high"
standardised_boston$crime[standardised_boston$crime=="(0.00739,9.92]"] <- "high"
# back to factor with correct levels
standardised_boston$crime<- factor(standardised_boston$crime, levels = c("low", "med_low", "med_high", "high"))

# Check head of edited table
head(standardised_boston)


# Build up test and train datasets

# choose randomly 80% of the rows
ind <- sample(nrow(standardised_boston),  size = n * 0.8)

# create train set
train <- standardised_boston[ind,]
dim(train)
# create test set
test <- standardised_boston[-ind,]
dim(test)

# Save n_cluster column
n_clusterit<- train$N_clusters

# remove original n_cluster from the dataset
train <- dplyr::select(train, -N_clusters)

# linear discriminant analysis
lda.fit <- lda(crime ~., data = train)

# print the lda.fit object
lda.fit

# Collect crime variables
target<- factor(train$crime)


# Collect predictor data
model_predictors <- dplyr::select(train, -crime)

# check the dimensions
dim(model_predictors)
dim(lda.fit$scaling)

# matrix multiplication
matrix_product <- as.matrix(model_predictors) %*% lda.fit$scaling
matrix_product <- as.data.frame(matrix_product)

# Next, access the plotly package. Create a 3D plot of the columns of the matrix.
library(plotly)
plot_ly(x = matrix_product$LD1, y = matrix_product$LD2, z = matrix_product$LD3, type= 'scatter3d', mode='markers', color=target)

# Let's colorcode clusters
plot_ly(x = matrix_product$LD1, y = matrix_product$LD2, z = matrix_product$LD3, type= 'scatter3d', mode='markers', color=n_clusterit)
```